---
title: "Edx_R_Linear Models in Practice"
author: "Joanna Yeh"
date: "2017/12/13"
output: html_document
---
```{r}
chooseCRANmirror(graphics=FALSE, ind=1)
knitr::opts_chunk$set(echo = TRUE)
```

HarvardX: PH525.2x
Introduction to Linear Models and Matrix Algebra
Section 3. Linear Models _ Practice and Exercise

The mouse diet example
* How to analyze the high fat diet data using linear models instead of directly applying a t-test. 

Read in the data and creat a quick stripchart:
```{r}
install.packages("lmtest", repos = "http://cran.us.r-project.org")
dat <- read.csv("femaleMiceWeights.csv") ##previously downloaded
filename <- "femaleMiceWeights.csv"
install.packages("downloader")
library(downloader)
if (!file.exists(filename)) {download(url, filename)}
```

```{r}
set.seed(1) #same jitter in stripchart
```
```{r}
dat <- read.csv("femaleMiceWeights.csv") ##previously downloaded
stripchart(dat$Bodyweight ~ dat$Diet, vertical=TRUE, method="jitter",
           main="Bodyweight over Diet")
```


```{r}
levels(dat$Diet) #factors have two levels, control diet and high fat diet
X <- model.matrix(~Diet, data = dat)
X #hf = 1, non-hf = 0
#colnames(X)
#指定 reference level and reassign to the column Diet
#dat$Diet <- relevel(dat$Diet, ref = "hf") 
#model.matrix(~Diet, data = dat) #chow = 1, non-chow = 0
```

Finally, we use our shortcut, lm, to run the linear model:
```{r}
fit <- lm(Bodyweight ~ Diet, data = dat)
summary(fit)
coefs <- coef(fit)
```

We can calculate this in R using our matrix multiplication operator %*%, the inverse function solve, and the transpose function t.

Method1. use linear model to get coefficients
```{r}
y <- dat$Bodyweight
X <- model.matrix(~Diet, data = dat)
solve(t(X) %*% X) %*% t(X) %*% y
```

Method 2. use split function (not linear model) to get the coefficients 
```{r}
s <- split(dat$Bodyweight, dat$Diet)
mean(s[["chow"]])
mean(s[["hf"]]~ mean(s[["chow"]]))
```

a visualization of the meaning of the coefficients with colored arrows:
```{r}
stripchart(dat$Bodyweight ~ dat$Diet, vertical=TRUE, method="jitter",
           main="Bodyweight over Diet", ylim=c(0,40), xlim=c(0,3))
a <- -0.25
lgth <- .1
library(RColorBrewer)
cols <- brewer.pal(3,"Dark2")
abline(h=0)
arrows(1+a,0,1+a,coefs[1],lwd=3,col=cols[1],length=lgth)
abline(h=coefs[1],col=cols[1])
arrows(2+a,coefs[1],2+a,coefs[1]+coefs[2],lwd=3,col=cols[2],length=lgth)
abline(h=coefs[1]+coefs[2],col=cols[2])
legend("right",names(coefs),fill=cols,cex=.75,bg="white")
```

Comparing simple two group to a t-test
```{r}
summary(fit)$coefficients
ttest <- t.test(s[["chow"]], s[["hf"]], var.equal=TRUE)
summary(fit)$coefficients[2,3]
ttest$statistic
#t.test(s[["hf"]], s[["chow"]], var.equal = TRUE)$statistics
```

#Practice 
1. You can make a design matrix X for a two group comparison either using model.matrix or simply with:

X = cbind(rep(1,nx + ny),rep(c(0,1),c(nx, ny)))

For a comparison of two groups, where the first group has nx=5 samples, and the second group has ny=7 samples, what is the element in the 1st row and 1st column of X^T X? What are all the other elements of (X^t X)?
```{r}
X <- cbind(rep(1,12),rep(c(0,1),c(5, 7)))
crossprod(X)
```


