---
title: "Interaction and Contrast"
author: "Joanna Yeh"
date: "2017/12/13"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Get the spider data from the scientific report
```{r}
url <- "https://raw.githubusercontent.com/genomicsclass/dagdata/master/inst/extdata/spider_wolff_gorb_2013.csv"
filename <- "spider_wolff_gorb_2013.csv"
library(downloader)
if (!file.exists(filename)) download(url, filename)
spider <- read.csv(filename, skip = 1) #skep = 1: skip the first line 
boxplot(spider$friction ~ spider$type + spider$leg, 
        col = c("grey90", "grey40"), las = 2,
        main = "Comparison of friction coefficients of different leg pairs")
```

Subset the dataset to just the row L1
```{r}
spider.sub <- spider[spider$leg == "L1",]
fit <- lm(friction ~ type, data = spider.sub)
summary(fit)
(coefs <- coef(fit))
s <- split(spider.sub$friction, spider.sub$type)
#show the coefficient 
mean(s[["pull"]])
mean(s[["push"]]) - mean(s[["pull"]])
```

Look into the model matrix
```{r}
X <- model.matrix(~type, data = spider.sub)
colnames(X)
head(X)
#rafalib package is the useful way to look at the moodel matrices
install.packages("rafalib")
library(rafalib)
imagemat(X, main = "Model matrix for LM with interactions")
```

#Examining the coefficients
```{r}
set.seed(1) #same jitter in stripchart
stripchart(split(spider.sub$friction, spider.sub$type), 
           vertical=TRUE, pch=1, method="jitter", las=2, xlim=c(0,3), ylim=c(0,2))
a <- -0.25
lgth <- .1
library(RColorBrewer)
cols <- brewer.pal(3,"Dark2")
abline(h=0)
arrows(1+a,0,1+a,coefs[1],lwd=3,col=cols[1],length=lgth)
abline(h=coefs[1],col=cols[1])
arrows(2+a,coefs[1],2+a,coefs[1]+coefs[2],lwd=3,col=cols[2],length=lgth)
abline(h=coefs[1]+coefs[2],col=cols[2])
legend("right",names(coefs),fill=cols,cex=.75,bg="white")
```

# A linear model with two variables
```{r}
X <- model.matrix(~ type + leg, data = spider)
colnames(X)
head(X)
library(rafalib)
imagemat(X, main = "Model matrix for linear model with 2 factors")
fit2 <- lm(friction ~ type + leg, data = spider)
summary(fit2)
(coefs <- coef(fit2))
```

# Arrow Plot
*arrows for each of the estimated coefficients in the model 
```{r}
spider$group <- factor(paste0(spider$leg, spider$type))
stripchart(split(spider$friction, spider$group), vertical = TRUE, 
           pch = 1, method = "jitter", las = 2, xlim = c(0,11), ylim = c(0,2))
a <- -0.25
lgth <- .1
library(RColorBrewer)
cols <- brewer.pal(5,"Dark2")
abline(h=0)
arrows(1+a,0,1+a,coefs[1],lwd=3,col=cols[1],length=lgth)
abline(h=coefs[1],col=cols[1])
arrows(3+a,coefs[1],3+a,coefs[1]+coefs[3],lwd=3,col=cols[3],length=lgth)
arrows(5+a,coefs[1],5+a,coefs[1]+coefs[4],lwd=3,col=cols[4],length=lgth)
arrows(7+a,coefs[1],7+a,coefs[1]+coefs[5],lwd=3,col=cols[5],length=lgth)
arrows(2+a,coefs[1],2+a,coefs[1]+coefs[2],lwd=3,col=cols[2],length=lgth)
segments(3+a,coefs[1]+coefs[3],4+a,coefs[1]+coefs[3],lwd=3,col=cols[3])
arrows(4+a,coefs[1]+coefs[3],4+a,coefs[1]+coefs[3]+coefs[2],lwd=3,col=cols[2],length=lgth)
segments(5+a,coefs[1]+coefs[4],6+a,coefs[1]+coefs[4],lwd=3,col=cols[4])
arrows(6+a,coefs[1]+coefs[4],6+a,coefs[1]+coefs[4]+coefs[2],lwd=3,col=cols[2],length=lgth)
segments(7+a,coefs[1]+coefs[5],8+a,coefs[1]+coefs[5],lwd=3,col=cols[5])
arrows(8+a,coefs[1]+coefs[5],8+a,coefs[1]+coefs[5]+coefs[2],lwd=3,col=cols[2],length=lgth)
legend("right",names(coefs),fill=cols,cex=.75,bg="white")

```

```{r}
s <- split(spider$friction, spider$group)
mean(s[["L1pull"]])
coefs[1]
mean(s[["L1push"]])
coefs[1] + coefs[2]
```

```{r}
means <- sapply(s, mean)
##the sample size of push or pull groups for each leg pair
ns <- sapply(s, length)[c(1,3,5,7)]
(w <- ns/sum(ns))
sum(w * (means[c(2,4,6,8)] - means[c(1,3,5,7)]))
coefs[2]
```

#Contrasting the coefficients

```{r}
library(contrast)
L3vsL2 <- contrast(fit2, list(leg = "L3", type = "pull"), list(leg = "L2", type = "pull"))
L3vsL2
coefs[4] - coefs[3]
(C <- L3vsL2$X )
L3vsL2.equiv <-  contrast(fit2, list(leg = "L3", type = "push"), list(leg = "L2", type = "push"))
L3vsL2.equiv 
```


#Contrast Exercises
Suppose we have an experiment with two species A and B, and two conditions: control and treated.
```{r}
species <- factor(c("A","A","B","B"))
condition <- factor(c("control","treated","control","treated"))
model.matrix(~ species + condition)
y <- rnorm(4)
fit <- lm( y ~ species + condition)
contrast(fit, list(species = "B", condition = "control"), list(species = "A", condition = "treated"))$X
```

#Contrast Exercises 2
```{r}
url <- "https://raw.githubusercontent.com/genomicsclass/dagdata/master/inst/extdata/spider_wolff_gorb_2013.csv"
filename <- "spider_wolff_gorb_2013.csv"
library(downloader)
if (!file.exists(filename)) download(url, filename)
spider <- read.csv(filename, skip=1)
fit2 <- lm(friction ~ type + leg, data = spider)
summary(fit2)
L4vsL2 <- contrast(fit2, list(leg = "L4", type = "pull"), list(leg = "L2", type = "pull"))
L4vsL2
```

```{r}
X <- model.matrix(~ type + leg, data=spider)
(Sigma <- sum(fit2$residuals^2)/(nrow(X) - ncol(X)) * solve(t(X) %*% X))
C <- matrix(c(0,0,-1,0,1),1,5)
Sigma[3,5]
```

#A linear model with interactions
```{r}
X <- model.matrix(~type + leg + type:leg, data = spider)
# equals to X <- model.matrix(~type*leg, data = spider)
colnames(X)
head(X)
imagemat(X, main = "Model matrix for linear model with interactions")
```

```{r}
fit3 <- lm(friction ~ type + leg + type:leg, data = spider)
summary(fit3)
coefs <- coef(fit3)
```

```{r}
spider$group <- factor(paste0(spider$leg, spider$type))
stripchart(split(spider$friction, spider$group), 
           vertical = TRUE, pch = 1, method = "jitter", las = 2, xlim = c(0,11), ylim = c(0,2))
cols <- brewer.pal(8, "Dark2")
abline(h=0)
arrows(1+a, 0, 1+a, coefs[1], lwd=3, col=cols[1], length = lgth)
abline(h=coefs[1], col = cols[1])
arrows(2+a, coefs[1], 2+a, coefs[1]+coefs[2], lwd = 3, col=cols[2], length = lgth)
arrows(3+a,coefs[1],3+a,coefs[1]+coefs[3],lwd=3,col=cols[3],length=lgth)
arrows(5+a,coefs[1],5+a,coefs[1]+coefs[4],lwd=3,col=cols[4],length=lgth)
arrows(7+a,coefs[1],7+a,coefs[1]+coefs[5],lwd=3,col=cols[5],length=lgth)

#now the interactions:
segments(3+a,coefs[1]+coefs[3],4+a,coefs[1]+coefs[3],lwd=3,col=cols[3])
arrows(4+a,coefs[1]+coefs[3],4+a,coefs[1]+coefs[3]+coefs[2],lwd=3,col=cols[2],length=lgth)
arrows(4+a,coefs[1]+coefs[2]+coefs[3],4+a,coefs[1]+coefs[2]+coefs[3]+coefs[6],lwd=3,col=cols[6],length=lgth)

segments(5+a,coefs[1]+coefs[4],6+a,coefs[1]+coefs[4],lwd=3,col=cols[4])
arrows(6+a,coefs[1]+coefs[4],6+a,coefs[1]+coefs[4]+coefs[2],lwd=3,col=cols[2],length=lgth)
arrows(6+a,coefs[1]+coefs[4]+coefs[2],6+a,coefs[1]+coefs[4]+coefs[2]+coefs[7],lwd=3,col=cols[7],length=lgth)

segments(7+a,coefs[1]+coefs[5],8+a,coefs[1]+coefs[5],lwd=3,col=cols[5])
arrows(8+a,coefs[1]+coefs[5],8+a,coefs[1]+coefs[5]+coefs[2],lwd=3,col=cols[2],length=lgth)
arrows(8+a,coefs[1]+coefs[5]+coefs[2],8+a,coefs[1]+coefs[5]+coefs[2]+coefs[8],lwd=3,col=cols[8],length=lgth)
legend("right",names(coefs),fill=cols,cex=.75,bg="white")
```

#Constrasts
```{r}
library(contrast)
L2push.vs.pull <- contrast(fit3, 
                           list(leg = "L2", type = "push"),
                           list(leg = "L2", type = "pull"))
L2push.vs.pull
C <- L2push.vs.pull$X
C
coef(fit3)
coefs[2] + coefs[6] ##we know this is also orange + yellow arrow
```

```{r}
library(multcomp) ##Available from CRAN
C <- matrix(c(0,0,0,0,0,-1,1,0), 1)
L3vsL2interaction <- glht(fit3, linfct=C) #glht = general linear hypotheses
summary(L3vsL2interaction)
coefs[7] - coefs[6] ##we know this is also brown - yellow
```

The question if the push vs. pull difference is different for different legs can be asked in general by the anova() function:
```{r}
anova(fit3)
```
#From the result as above, P value of F test is really small, so, we can interpret that the difference we see across leg pair in push vs pull are more than we expected by chance. Hence, the push vs pull is significantly difference across leg pairs when looking at all four leg pairs.

#Interactions Exercises
```{r}
#loading the spider dataset
url <- "https://raw.githubusercontent.com/genomicsclass/dagdata/master/inst/extdata/spider_wolff_gorb_2013.csv"
filename <- "spider_wolff_gorb_2013.csv"
library(downloader)
if (!file.exists(filename)) download(url, filename)
spider <- read.csv(filename, skip=1)
```

```{r}
spider$log2friction <- log2(spider$friction)
boxplot(log2friction ~ type*leg, data=spider)
fit4 <- lm(log2friction ~ type + leg + type:leg, data = spider)
summary(fit4)
anova(fit4)
contrast(fit4, list(type="pull",leg="L2"), list(type="pull",leg="L1"))
coef(fit4)["legL2"]
contrast(fit4, list(type="push",leg="L2"), list(type="push",leg="L1"))
coef(fit4)["legL2"] + coef(fit4)["typepush:legL2"]
```

#5
```{r}
N <- 40
p <- 4
group <- factor(rep(1:p,each=N/p))
X <- model.matrix(~ group)
Y <- rnorm(N,mean=42,7)
mu0 <- mean(Y)
initial.ss <- sum((Y - mu0)^2)
s <- split(Y, group)
after.group.ss <- sum(sapply(s, function(x) sum((x - mean(x))^2))) #the residual sum of squares
(group.ss <- initial.ss - after.group.ss) #the explanatory power of the group variable
group.ms <- group.ss / (p - 1)
after.group.ms <- after.group.ss / (N - p)
f.value <- group.ms / after.group.ms
```

```{r}
N <- 40
p <- 4
group <- factor(rep(1:p,each=N/p))
X <- model.matrix(~ group)
num_random <- 1000
set.seed(1)
F_value <- replicate(num_random, {
  Y <- rnorm(N,mean=42,7)
  mu0 <- mean(Y)
  initial.ss <- sum((Y - mu0)^2)
  s <- split(Y, group)
  after.group.ss <- sum(sapply(s, function(x) sum((x - mean(x))^2))) #the residual sum of squares
  (group.ss <- initial.ss - after.group.ss) #the explanatory power of the group variable
  group.ms <- group.ss / (p - 1)
  after.group.ms <- after.group.ss / (N - p)
  f.value <- group.ms / after.group.ms
  return(f.value)
})
mean(F_value)
# calculate the F-value for 1000 random versions of Y
```

```{r}
hist(F_value, col="grey", border="white", breaks=50, freq=FALSE)
xs <- seq(from=0,to=6,length=100)
lines(xs, df(xs, df1 = p - 1, df2 = N - p), col="red")
```

```{r}
spider$group <- factor(paste0(spider$leg, spider$type))
X <- model.matrix(~ 0 + group, data = spider)
colnames(X)
head(X)
imagemat(X, main = "Model matrix for linear model with interaction")
fit5 <- lm(friction ~ 0 + group, data = spider)
summary(fit5)
coefs <- coef(fit5)
```

#Examining the coefficients
```{r}
stripchart(split(spider$friction, spider$group),
           vertical = TRUE, pch = 1, method = "jitter", las = 2,
           xlim = c(0,11), ylim = c(0,2))
cols <- brewer.pal(8, "Dark2")
abline(h=0)
for(i in 1:8){
  arrows(i+a, 0, i+a, coefs[i], lwd = 3, col = cols[i], length = lgth)
}
legend("right", names(coefs), fill = cols, cex = .75, bg = "white")
```

#Simple contrasts using the contrast package
```{r}
groupL2push.vs.pull <- contrast(fit5,
                                list(group = "L2push"),
                                list(group = "L2pull"))
groupL2push.vs.pull
coefs[4] - coefs[3]
```

#Differences of differences when there is not an intercept
```{r}
C <- matrix(c(0, 0, 1, -1, -1, 1, 0, 0), 1)
groupL3.vs.L2interaction <- glht(fit5, linfct = C)
summary(groupL3.vs.L2interaction)
(coefs[6] - coefs[5]) - (coefs[4] - coefs[3])
```

